---
title: "4_Spatial Analysis"
author: "Nadia Kennar, Research Associate with the UKDS"
date: "10th of March 2026"
output:
  html_document:
    df_print: paged
  html_notebook: default
toc: yes 
---

```{r set up, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Intro 

In this final section I will 

1. Filling missing LSOA counts using neighbors.

2. Heatmaps / interactive maps of ASB.

3. Time series of ASB.

## Load packages

As always the first step is to load the necessary R packages via the library function. If you do not have these packages installed then please follow the instructions in the *Preliminary Task.Rmd* file. 

## Load Packages

```{r setup}

# for data reading/manipulation 
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(janitor)
library(readxl)
# for spatial data and gis
library(sf)
library(ggmap)
library(ggplot2)
library(ggspatial)
library(spdep)
library(leaflet) 
library(RColorBrewer)
library(tmap)

```


## Spatial Interpolation 

We’ll look at a simple way to estimate missing crime counts for LSOAs by borrowing information from neighbouring areas. This is not full-blown statistical interpolation, but a pragmatic way to avoid ‘holes’ in our maps, using the idea that neighboring areas tend to be similar.

Spatial interpolation is a statistical technique used to predict or estimate values at unsampled locations within an area covered by existing observations. It's based on the principle that spatially or temporally close points tend to have similar values—a concept known as spatial autocorrelation. In the context of crime data, spatial interpolation can be particularly useful for several reasons:

*1. Fill Gaps: Estimate crime rates in areas where data are missing or not collected, providing a more complete spatial representation of crime.
*2. Smooth Crime Rates: Generate continuous surfaces of crime rates or risk across a geographic area, helping to identify hotspots and patterns that might not be evident from discrete data points alone.
*3. Resource Allocation: Inform law enforcement and public safety officials about potential areas of increased crime activity, aiding in strategic planning and resource allocation.
*4. Public Information: Provide communities with detailed maps of crime risk, enhancing awareness and preventive measures.

Lets explore our data. 

```{r}
# crime <- read_csv("/Users/user/Documents/Mapping_Crime_Data_R_2025/Data/crime_data/2024-02/2024-02-surrey-street.csv") %>% janitor::clean_names()
# 
# summary(crime)

url <- "https://raw.githubusercontent.com/UKDataServiceOpen/Mapping_Crime_Data_R_2026/main/data/crime_data/2025-02/2025-02-surrey-street.csv"
crime <- read_csv(url) %>% janitor::clean_names()

```


We know that there are 5970 observations overall, but what about just for Surrey Heath which is a boundary area within the county of Surrey. 

We can use filter only those crimes that have taken place in Surrey Heath by using the dplyr and stringr packages

```{r}
library(stringr)
surrey_heath_crimes <- crime %>%
  filter(str_detect(lsoa_name, "^Surrey Heath"))

# Explore the filtered dataset
head(surrey_heath_crimes)
summary(surrey_heath_crimes)
```

This code filters the original crime dataset to keep only the records associated with LSOAs in the Surrey Heath area, based on the naming convention in the lsoa_name column, and stores this subset of data in a new variable for further analysis or operations.

We can now see that there are 370 observations, that is individual crime counts, across the boundary area of Surrey Heath. 

We know need to group these individual crime counts by lsoas in order to obtain the aggregated statistics. This has been repeated in multiple sections so I won't expand too much here. We will create this into a new object named 'cleaned_crimes_grouped_by_lsoas' 


```{r}
cleaned_crimes_grouped_by_lsoa <- surrey_heath_crimes %>%
  group_by(lsoa_code) %>%
  summarise(count = n())

# View the count of crimes by type
cleaned_crimes_grouped_by_lsoa
```


So what can we see? Well there are now 54 observations, in this case representing the number of LSOAs in Surrey Heath. However, if we compare this to our Shapefile of Surrey Heath, there are 55 observations which tells us there is 1 missing LSOAs. 

We can run a spatial interpolation to account for this by estimating the missing crime counts based on the average of neighboring LSOAs. 

Lets first identify the missing LSOA in the crime data set. We can use the setdiff() function from base R to do so...

```{r}
missing_lsoas <- setdiff(shp_file$lsoa21cd, cleaned_crimes_grouped_by_lsoa$lsoa_code)
```

There we go! We have identified the missing LSOA is "E01030778"

In order to now run the spatial interpolation on the average of neighboring LSOAs, spatial visualization is required in order to understand if these LSOA does in fact have any neighboring LSOAs to establish a mean. This is because if the LSOA does not have any neighboring areas, we will not be able to establish a mean and therefore estimate a crime count value.

We can first establish a list of neighbors where each element corresponds to a LSOA within the shapefile (shp_file) and contains the indices of neighboring LSOAS.

```{r}
neighbors <- st_touches(shp_file, shp_file)
```

This allows us to now identify the LSOAs with no neighbors. This line uses sapply() to apply a function to each element of the neighbors list. The function checks if the length of each element (list of neighbor indices) is 0, which would indicate that an LSOA has no neighbors.

```{r}
# Identify LSOAs with no neighbors
no_neighbors <- sapply(neighbors, function(x) length(x) == 0)
```


We then want to extract the indices (or IDs) of LSOAs with no neighbors and then extract the corresponding LSOAS. 

```{r}
# Extract the indices or IDs of LSOAs with no neighbors
lsoas_no_neighbors <- which(no_neighbors)

# Extract the corresponding LSOA codes or IDs
lsoas_with_no_neighbors <- shp_file$lsoa_code[lsoas_no_neighbors]
```


If lsoas_with_no_neighbors is empty, all LSOAs have at least one neighbor. If lsoas_with_no_neighbors contains LSOA codes or IDs, these are the LSOAs with no neighboring LSOAs within your dataset.

This would mean that we know that the LSOA does in fact have neighboring areas and therefore we can run some calculations on our neighboring areas to interpolate that missing value. For each missing LSOA, we need to find neighboring LSOAs and calculate the average crime count. 

This function iterates over a list of LSOAs (missing_lsoas) that lack crime data. For each missing LSOA, it:

*Identifies neighboring LSOAs that share a boundary with it.
*Gathers crime counts from these neighboring LSOAs.
*Calculates the average crime count from these neighbors.
*Rounds this average to the nearest whole number.
*Prints out this average for verification.
*If the average crime count is not calculated properly (e.g., it's NA or 0), it sets a default value of 1.
*Finally, it adds this calculated average crime count back into the dataset for the missing LSOA, effectively estimating the crime count for areas where it was originally missing

This process helps fill in gaps in the dataset, ensuring each LSOA has a crime count value, which is crucial for comprehensive spatial analysis.

```{r}

for(missing_lsoa in missing_lsoas) {
    neighbors_indices <- st_touches(shp_file[shp_file$lsoa21cd == missing_lsoa, , drop = FALSE], shp_file)[[1]]
    neighbor_lsoas <- shp_file$lsoa21cd[neighbors_indices]

    # Extracting crime counts for neighboring LSOAs
    neighbor_crime_counts <- cleaned_crimes_grouped_by_lsoa %>% 
                             filter(lsoa_code %in% neighbor_lsoas) %>%
                             pull(count) # Use pull to extract the 'count' column as a vector

    # Calculating the average crime count
    avg_crime_count <- mean(neighbor_crime_counts, na.rm = TRUE)
    
    # Round the average to the nearest whole number
    avg_crime_count_rounded <- round(avg_crime_count)

    # Debug: print the average crime count to verify
    print(paste("Average crime count for missing LSOA", missing_lsoa, ":", avg_crime_count))

    # Check if avg_crime_count is not calculated properly
    if(is.na(avg_crime_count) || avg_crime_count == 0) {
      # Apply a different strategy or set a default value other than 0 if that makes sense in your context
      avg_crime_count <- 1 # Example: setting a default value if mean calculation fails
    }

    # Update the dataset with the calculated average for the missing LSOA
    cleaned_crimes_grouped_by_lsoa <- rbind(cleaned_crimes_grouped_by_lsoa, data.frame(lsoa_code = missing_lsoa, count = avg_crime_count_rounded))
}


class(cleaned_crimes_grouped_by_lsoa)
```


Did you notice how the object 'cleaned_crimes_grouped_by_lsoas' has now changed.

```{r}
cleaned_crimes_grouped_by_lsoa[55, ]
```


Lets compare the previous surrey_lsoa (with the misssing lsoa data) and the surrey_lsoa_new (no missing lsoa data). Lets look at how these might be different both statistically and visually: 

```{r}
# Ensure both datasets are spatially joined to the original shapefile
surrey_lsoa <- left_join(shp_file, crimes_grouped_by_lsoa, by = c("lsoa21cd" = "lsoa_code"))
surrey_lsoa_new <- left_join(shp_file, cleaned_crimes_grouped_by_lsoa, by = c("lsoa21cd" = "lsoa_code"))
```



```{r}
# Statistically Compare 
mean_original <- mean(surrey_lsoa$count, na.rm = TRUE)
mean_imputed <- mean(surrey_lsoa_new$count)

median_original <- median(surrey_lsoa$count, na.rm = TRUE)
median_imputed <- median(surrey_lsoa_new$count)

# Print the results
cat("Mean Crime Count - Original Data (with missing):", mean_original, "\n")
cat("Mean Crime Count - Imputed Data (no missing):", mean_imputed, "\n\n")

cat("Median Crime Count - Original Data (with missing):", median_original, "\n")
cat("Median Crime Count - Imputed Data (no missing):", median_imputed, "\n")
```


```{r}
# Visually Compare
# Plotting the original dataset with missing value
ggplot() +
  geom_sf(data = surrey_lsoa, aes(fill = count), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey", guide = guide_legend(title = "Crime Count")) +
  ggtitle("Original Data with Missing LSOA Crime Count") +
  theme_minimal()

# Plotting the dataset with imputed value
ggplot() +
  geom_sf(data = surrey_lsoa_new, aes(fill = count), color = NA) +
  scale_fill_viridis_c(option = "plasma", guide = guide_legend(title = "Crime Count")) +
  ggtitle("Imputed Data with No Missing LSOA Crime Count") +
  theme_minimal()


```

# Moran's I 

We’ve implicitly used the idea of spatial autocorrelation throughout: nearby areas tend to have similar crime levels. That’s why borrowing information from neighbouring LSOAs made sense in our ‘interpolation’ step. In more advanced work, we can measure this formally using statistics like Moran’s I and build spatial regression models, but today we’re just using the concept to guide simple, transparent methods. 

Here is a very quick example of using Moran's I

```{r}
library(spdep)

# build neighbours from polygons
nb <- poly2nb(surrey_lsoa_new)
lw <- nb2listw(nb, style = "W")

moran.test(surrey_lsoa_new$count, lw)

```

Then verbally: If Moran’s I is positive and significant, it confirms that high‑crime areas are clustered.



## Heat Maps (Chloropleth maps)

As I'm specifically interested in creating heatmaps of ASB across surrey heath, there is a little bit of data preparation needed. The steps are just repeated from above but they can be summarised as such:

*1) filter for asb within the crime dataframe (and within surrey heath, although this is not necessary as the shapefile is from surrey heath, but i want to impute some missing values before hand) 
*2) group by lsoa and summarise the counts
*3) find the missing lsoas
*4) impute the missing values for the lsoas
*5) left_join to the shapefile


```{r}
# Filter for ASB within Surry Heath
asb_data <- crime %>%
  filter(crime_type == "Anti-social behaviour") %>% 
  filter(str_detect(lsoa_name, "^Surrey Heath"))

# Group by LSOA and summarise the counts
asb_grouped_by_lsoa <- asb_data %>%
  group_by(lsoa_code) %>%
  summarise(count=n())

# Find the missing LSOAs
missing_lsoas <- setdiff(shp_file$lsoa21cd, asb_grouped_by_lsoa$lsoa_code)

# Impute the missing values 
for(missing_lsoa in missing_lsoas) {
    neighbors_indices <- st_touches(shp_file[shp_file$lsoa21cd == missing_lsoa, , drop = FALSE], shp_file)[[1]]
    neighbor_lsoas <- shp_file$lsoa21cd[neighbors_indices]

    # Extracting crime counts for neighboring LSOAs
    neighbor_crime_counts <- asb_grouped_by_lsoa %>% 
                             filter(lsoa_code %in% neighbor_lsoas) %>%
                             pull(count) # Use pull to extract the 'count' column as a vector

    # Calculating the average crime count
    avg_crime_count <- mean(neighbor_crime_counts, na.rm = TRUE)
    
    # Round the average to the nearest whole number
    avg_crime_count_rounded <- round(avg_crime_count)

    # Debug: print the average crime count to verify
    print(paste("Average crime count for missing LSOA", missing_lsoa, ":", avg_crime_count))

    # Check if avg_crime_count is not calculated properly
    if(is.na(avg_crime_count) || avg_crime_count == 0) {
      # Apply a different strategy or set a default value other than 0 if that makes sense in your context
      avg_crime_count <- 1 # Example: setting a default value if mean calculation fails
    }

    # Update the dataset with the calculated average for the missing LSOA
    asb_grouped_by_lsoa <- rbind(asb_grouped_by_lsoa, data.frame(lsoa_code = missing_lsoa, count = avg_crime_count_rounded))
}

# Left_join the now cleaned data back into the shapefile
asb_lsoa <- left_join(shp_file, asb_grouped_by_lsoa, by = c("lsoa21cd" = "lsoa_code"))
```



### Using GGPPLOT2

This ggplot code snippet creates a heat map visualizing the distribution of ASB incidents across LSOAs. The use of the magma color palette from viridis ensures that the visualization is both attractive and perceptually accurate, enabling viewers to easily see where ASB incidents are more or less concentrated. 

```{r}

ggplot(data = asb_lsoa) +
  geom_sf(aes(fill = count), color = NA) + # geom_sf plots the spatial data, fill controls the color based on ASB count
  scale_fill_viridis_c(option = "magma", direction = -1, na.value = "white", 
                       guide = guide_colorbar(title = "ASB Count")) + # Adjusts the color scale
  labs(title = "ASB Incidents by LSOA", 
       subtitle = "Visualised using ggplot2") + 
  theme_minimal()

```


### Using leaflet


This code snippet creates an interactive map using the leaflet package in R to visualize Anti-Social Behaviour (ASB) incidents across Lower Super Output Areas (LSOAs) based on the asb_lsoa dataset. The dataset is expected to be an sf object containing geometries for each LSOA and a count variable representing the number of ASB incidents in each area. 

Note: Leaflet wants geometries in WGS84 (EPSG 4326), so we transform before passing the layer in. 

```{r}
#asb_lsoa <- st_transform(asb_lsoa, 4326)
asb_lsoa <- st_transform(asb_lsoa, 4326)

# Define a color palette function based on the 'count' column
colorPalette <- colorNumeric(palette = "viridis", domain = asb_lsoa$count, na.color = "transparent")

# Create the Leaflet map
leaflet(asb_lsoa) %>%
  addProviderTiles(providers$CartoDB.Positron) %>% # Adds a light-themed base map
  addPolygons(
    fillColor = ~colorPalette(count), # Apply the color based on the 'count' column
    weight = 1, # Border weight
    opacity = 1, # Border opacity
    color = "white", # Border color
    fillOpacity = 0.7, # Fill opacity
    popup = ~paste0("LSOA: ", lsoa21cd, "<br>Count: ", count) # Popup content
  ) %>%
  addLegend("bottomright", pal = colorPalette, values = ~count,
            title = "ASB Counts",
            opacity = 1)



```



## Time Series Analysis (optional activity)

After completing our spatial interpolation, we now shift our focus to analyzing trends over time for Anti-Social Behaviour (ASB) incidents across Surrey Heath.

Crime patterns are rarely static, and time series analysis helps us:
✔ Identify trends – Are ASB incidents increasing or decreasing?
✔ Detect seasonal patterns – Do incidents peak in certain months?
✔ Compare variations across LSOAs – Are some areas more volatile than others?

What We Will Do
In this section, we will:
1️⃣ Load and process 12 months of ASB data dynamically.
2️⃣ Visualize ASB trends per LSOA (messy but insightful).
3️⃣ Aggregate ASB trends across all LSOAs for a clearer picture.
4️⃣ Apply a rolling average to smooth fluctuations and highlight long-term trends.

By the end of this section, we will have a comprehensive understanding of how ASB incidents evolve over time and how we can use temporal patterns to inform crime prevention strategies.


### Load & Process ASB Crime Data  
We dynamically load **12 months of crime data** from the `Data/crime_data` directory.

```{r load-data}
# Define path to the crime data folder
#crime_data_path <- "Data/crime_data"
# crime_data_path <- "/Users/user/Documents/Mapping_Crime_Data_R_2025/Data/crime_data"
# 
# # List all monthly folders dynamically (last 12 months)
# monthly_folders <- dir_ls(crime_data_path, type = "directory") %>% tail(12)
# 
# # Function to read each month's crime data
# read_crime_data <- function(folder) {
#   file_path <- dir_ls(folder, regexp = "surrey-street.csv$")
#   read_csv(file_path) %>%
#     clean_names() %>%
#     mutate(month = basename(folder)) # Extract folder name as new column
# }
# 
# # Read all crime data from the last 12 months
# all_crime_data <- map_dfr(monthly_folders, read_crime_data)

##
library(tidyverse)
library(fs)
library(janitor)

# Define base GitHub URL
github_base <- "https://raw.githubusercontent.com/UKDataServiceOpen/Mapping_Crime_Data_R_2026/main/data/crime_data"

# List last 12 months (update as needed)
months <- format(seq(Sys.Date(), by = "-1 month", length.out = 12), "%Y-%m")

# Function to download and read each month's crime data
read_crime_data <- function(month) {
  file_url <- paste0(github_base, "/", month, "/", month, "-surrey-street.csv")
  
  # Temporary file to store downloaded data
  temp_file <- tempfile(fileext = ".csv")
  
  # Download the file
  download.file(file_url, temp_file, mode = "wb")
  
  # Read the CSV file
  read_csv(temp_file) %>%
    clean_names() %>%
    mutate(month = month) # Add month as a column
}

# Read and merge all crime data from the last 12 months
all_crime_data <- map_dfr(months, possibly(read_crime_data, NULL))

# View dataset
head(all_crime_data)


```

---

### Filter & Aggregate ASB Crimes  
We filter for **Anti-Social Behaviour** incidents in **Surrey Heath** and aggregate by **LSOA & Month**.

```{r filter-aggregate}
# Filter for ASB crimes in Surrey Heath
asb_crime_data <- all_crime_data %>%
  filter(crime_type == "Anti-social behaviour") %>%
  filter(str_detect(lsoa_name, "^Surrey Heath"))

# Aggregate by LSOA and Month
asb_grouped_by_lsoa <- asb_crime_data %>%
  mutate(year = substr(month, 1, 4), month = substr(month, 6, 7)) %>%
  group_by(lsoa_code, year, month) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(lsoa_code, year, month)

# Create a date column for time series plotting
asb_grouped_by_lsoa <- asb_grouped_by_lsoa %>%
  mutate(date = as.Date(paste(year, month, "01", sep = "-")))
```

---
 
### **1. Individual LSOA ASB Trends** (*Messy but insightful*)  
Each line represents a different LSOA.

```{r plot-lsoa-trends}
# Plot individual LSOA trends
ggplot(asb_grouped_by_lsoa, aes(x = date, y = count, group = lsoa_code)) +
  geom_line(alpha = 0.3, color = "gray") +  # Faint individual lines
  geom_smooth(method = "loess", se = FALSE, color = "blue", size = 1.2) +  # Trend line
  labs(title = "ASB Trends Over Time (Per LSOA)",
       x = "Month", y = "Number of ASB Incidents") +
  theme_minimal()
```

---

### **2. Aggregated ASB Trends** (*Cleaner and more readable*)  
We sum ASB counts across **all LSOAs** per month.

```{r aggregate-trend}
# Aggregate ASB count by month
asb_trend <- asb_grouped_by_lsoa %>%
  group_by(date) %>%
  summarise(total_count = sum(count), .groups = "drop")

# Plot aggregated trend
ggplot(asb_trend, aes(x = date, y = total_count)) +
  geom_line(color = "blue", size = 1.2) +  # Trend line
  geom_point(color = "red", size = 2) +  # Highlight monthly points
  labs(title = "Aggregated ASB Trends Over Time",
       x = "Month", y = "Total ASB Incidents") +
  theme_minimal()
```

---

### **3. Smoothed ASB Trend (Rolling Average)**  
We apply a **3-month rolling average** to smooth fluctuations.

```{r rolling-average}
# Add 3-month rolling average
library(zoo)
asb_trend <- asb_trend %>%
  mutate(rolling_avg = rollmean(total_count, k = 3, fill = NA, align = "right"))

# Plot rolling average trend
ggplot(asb_trend, aes(x = date)) +
  geom_line(aes(y = total_count), color = "lightblue", size = 1) +  # Raw data
  geom_line(aes(y = rolling_avg), color = "blue", size = 1.5) +  # Smoothed trend
  geom_point(aes(y = total_count), color = "red", size = 2) +
  labs(title = "Aggregated ASB Trends with Rolling Average",
       x = "Month", y = "Total ASB Incidents") +
  theme_minimal()
```








## Activity 

Compare the heat maps now using the ASB data WITH the missing values. Follow the steps below by first preparing your dataset and then making the plots. 

*1) Filter for just ASB within Surrey Heath 

```{r}
asb_data2 <- crime %>%
  filter(.... == "Anti-social behaviour") %>% 
  filter(str_detect(...., "^Surrey Heath"))
```

*2) Group by LSOA and summarise the crime count

```{r}
asb_grouped_by_lsoa2 <- asb_data %>%
  group_by(.....) %>%
  summarise(.....=n())
```

*3) Left_join the new grouped ASB dataset to the shapefile

Remember we are skipping the imputation method here as we want to see the affect of including missing values into our maps. So go ahead and simply join the grouped data to the shapefile

```{r}
asb_lsoa_missing <- left_join(...., ...., by = c("lsoa21cd" = "lsoa_code"))
```

*4) Now plot with ggplot 

```{r}
ggplot(data = ...) +
  geom_sf(aes(fill = ...), color = NA) + 
  scale_fill_viridis_c(option = "magma", direction = -1, na.value = "white", 
                       guide = guide_colorbar(title = "ASB Count")) + # Adjusts the color scale
  labs(title = "ASB Incidents by LSOA", 
       subtitle = "Visualised using ggplot2") + 
  theme_minimal()


```

*5) Now plot with leaflet 

```{r}
# Define a color palette function based on the 'count' column
colorPalette <- colorNumeric(palette = "viridis", domain = asb_lsoa_missing$count, na.color = "transparent")

# remember to first transform the object into the correct CRS
asb_lsoa_missing <- .....(asb_lsoa_missing, ....)

# Create the Leaflet map
leaflet(....) %>%
  addProviderTiles(providers$CartoDB.Positron) %>% # Adds a light-themed base map
  addPolygons(
    fillColor = ~colorPalette(count), 
    weight = 1, # Border weight
    opacity = 1, # Border opacity
    color = "white", # Border color
    fillOpacity = 0.7, # Fill opacity
    popup = ~paste0("LSOA: ", lsoa21cd, "<br>Count: ", count) 
  ) %>%
  addLegend("bottomright", pal = colorPalette, values = ~count,
            title = "ASB Counts",
            opacity = 1)

```



Data Citation: 

1.	UK Police API. (2026). Street-level crime data. data.police.uk. Retrieved February 27, 2026, from https://data.police.uk
2.	UK Data Service Census Support. (2021). England LSOA boundaries 2021. Retrieved from https://borders.ukdataservice.ac.uk 
3.	Office for National Statistics. (2022). Census 2021 Table TS001: Number of usual residents. Nomis/UK Data Service. Retrieved from https://statistics.ukdataservice.ac.uk
4.	Office for National Statistics. (2022). Census 2021 Table TS066: Economic acitivty. Nomis/UK Data Service. Retrieved from https://statistics.ukdataservice.ac.uk
5.	Office for National Statistics. (2021). Rural Urban Classification (2021) of LSOAs in England and Wales. Open Geography Portal. Retrieved from https://geoportal.statistics.gov.uk 


