---
title: "Activity Solutions Sections 1-4"
author: "Nadia Kennar, Research Associate with the UKDS"
date: "10th of March 2026"
output:
  html_notebook: default
toc: yes
---


## Solution to Activty 1

```{r}
subset(sf, crime_type == "Anti-social behaviour") 

drugs <- subset(sf, crime_type == "Drugs") %>%  
  select(-c(1, 9, 10))      #this line is not necessary but helps to neaten your data as we are removing the columns that are not of interest

ggplot() +
  annotation_map_tile() +
  geom_sf(data = drugs) 

```


## Solution to Activity 2 

```{r}
#1)

h <- tm_shape(surrey_lsoa) + 
  tm_fill("count", style = "bclust") + 
  tm_borders(alpha = 0.3)


b <- tm_shape(surrey_lsoa) + 
  tm_fill("count", style = "hclust") + 
  tm_borders(alpha = 0.3)


#2) 

tmap_arrange(h, b)

#3) 

tmap_mode("view")


tm_shape(surrey_lsoa) + 
  tm_fill("count", style = "bclust") + 
  tm_borders(alpha = 0.3)


```


## Solution to Activity 3
```{r}

# 1) Calculate crime rate relative to economically inactive population
surrey_lsoa <- surrey_lsoa %>% 
  mutate(crime_rate2 = (count / econ_inactive) * 1000)

# 2) Plot using ggplot
ggplot() + 
  annotation_map_tile() + 
  geom_sf(data = surrey_lsoa, aes(fill = crime_rate2), alpha = 0.5) + 
  scale_fill_gradient2(name = "Crime Rate \n(per 1,000 economically inactive)")

# Improved ggplot version
ggplot(surrey_lsoa) + 
  geom_sf(aes(fill = crime_rate2), color = "black", lwd = 0.2) +
  scale_fill_viridis_c(option = "magma", name = "Crime Rate\n(per 1,000 economically inactive)") +
  labs(title = "Crime Rate by LSOA in Surrey",
       subtitle = "Relative to economically inactive population",
       caption = "Source: Police.uk & Census 2021") +
  theme_minimal()

# 3) Plot using tmap
tm_shape(surrey_lsoa) + 
  tm_fill("crime_rate2", style = "quantile",
          title = "Crime Rate\n(per 1,000 econ. inactive)") + 
  tm_borders(alpha = 0.3)

# 4) Compare crime rate vs residential population side by side
e <- tm_shape(surrey_lsoa) + 
  tm_fill("crime_rate2", 
          style = "quantile", 
          title = "Econ. Inactive Pop") + 
  tm_borders(alpha = 0.3) +
  tm_layout(main.title = "Crime Rate: Econ. Inactive")

f <- tm_shape(surrey_lsoa) + 
  tm_fill("crime_rate", 
          style = "quantile", 
          title = "Residential Pop") + 
  tm_borders(alpha = 0.3) +
  tm_layout(main.title = "Crime Rate: Residential")

tmap_mode("plot")
tmap_arrange(e, f)

# 5) Economic inactivity crime rate by urban/rural (multivariate!)
tmap_mode("view")
tm_shape(surrey_lsoa) +
  tm_fill("crime_rate2", style = "jenks", palette = "Blues",
          title = "Crime Rate\n(per 1,000 econ. inactive)") +
  tm_facets(by = "urban_rural_flag") +
  tm_borders("black", lwd = 0.3) +
  tm_layout(title = "Crime Rates by Economic Inactivity: Urban vs Rural")
```


## Solution to Activity 4 


```{r}
## Activity 

Compare the heat maps now using the ASB data WITH the missing values. Follow the steps below by first preparing your dataset and then making the plots. 

*1) Filter for just ASB within Surrey Heath 

```{r}
asb_data2 <- crime %>%
  filter(crime_type == "Anti-social behaviour") %>% 
  filter(str_detect(losa_name, "^Surrey Heath"))
```

*2) Group by LSOA and summarise the crime count

```{r}
asb_grouped_by_lsoa2 <- asb_data %>%
  group_by(lsoa_code) %>%
  summarise(count =n())
```

*3) Left_join the new grouped ASB dataset to the shapefile

Remember we are skipping the imputation method here as we want to see the affect of including missing values into our maps. So go ahead and simply join the grouped data to the shapefile

```{r}
asb_lsoa_missing <- left_join(shp_file, asb_grouped_by_lsoa2, by = c("lsoa21cd" = "lsoa_code"))
```

*4) Now plot with ggplot 

```{r}
ggplot(data = asb_lsoa_missing) +
  geom_sf(aes(fill = count), color = NA) + 
  scale_fill_viridis_c(option = "magma", direction = -1, na.value = "white", 
                       guide = guide_colorbar(title = "ASB Count")) + # Adjusts the color scale
  labs(title = "ASB Incidents by LSOA", 
       subtitle = "Visualised using ggplot2") + 
  theme_minimal()


```

*5) Now plot with leaflet 

```{r}
# Define a color palette function based on the 'count' column
colorPalette <- colorNumeric(palette = "viridis", domain = asb_lsoa_missing$count, na.color = "transparent")

# remember to first transform the object into the correct CRS
asb_lsoa_missing <- st_transform(asb_lsoa_missing, 4326)

# Create the Leaflet map
leaflet(asb_lsoa_missing) %>%
  addProviderTiles(providers$CartoDB.Positron) %>% # Adds a light-themed base map
  addPolygons(
    fillColor = ~colorPalette(count), 
    weight = 1, # Border weight
    opacity = 1, # Border opacity
    color = "white", # Border color
    fillOpacity = 0.7, # Fill opacity
    popup = ~paste0("LSOA: ", lsoa21cd, "<br>Count: ", count) 
  ) %>%
  addLegend("bottomright", pal = colorPalette, values = ~count,
            title = "ASB Counts",
            opacity = 1)

```




Data Citation: 

1.	UK Police API. (2026). Street-level crime data. data.police.uk. Retrieved February 27, 2026, from https://data.police.uk
2.	UK Data Service Census Support. (2021). England LSOA boundaries 2021. Retrieved from https://borders.ukdataservice.ac.uk 
3.	Office for National Statistics. (2022). Census 2021 Table TS001: Number of usual residents. Nomis/UK Data Service. Retrieved from https://statistics.ukdataservice.ac.uk
4.	Office for National Statistics. (2022). Census 2021 Table TS066: Economic acitivty. Nomis/UK Data Service. Retrieved from https://statistics.ukdataservice.ac.uk
5.	Office for National Statistics. (2021). Rural Urban Classification (2021) of LSOAs in England and Wales. Open Geography Portal. Retrieved from https://geoportal.statistics.gov.uk 

